---
title: "Cleaning up the data"
output:
  html_notebook: default
  pdf_document: default
---

Data was collected from the HUD Point-In-Time estimates of homelessness assessment reports for 2007 through 2017. 

Data was compiled and queried within data.world and the data clean-up was conducted in R studio. The focus was on organizing the data in such a way that it would be easy to group by region, by state or by year. I also wanted to be able to combine data on the number of shelter beds available for homeless individuals per state with the number of homeless individuals per state. 

SQL was used to pull data from the various sheets into one dataframe within R. The data was then organized, cleaned and joined to create one dataframe. 

Dataframe clean-up included:
* Removing blanks
* Spelling out state names
* Removing US territories that will not be used in analysis
* Removing rows of data with totals 
* Breaking down data according to regions
* Combining data points from shelter beds dataframe and homeless population dataframe

### Steps for Compiling and Cleaning up the Data 

Libraries used
```{r, echo =TRUE, results='hide'}
library(data.world) #used to connect to data.world 
library(dplyr) #to manipulate and transform the data
```

```{r, echo =TRUE, results='hide'}
load(".RData")
```
To pull data.world you first have to configure the system so that R studio can directly communicate with data.world. Data.world provides each user with an API token which is used to integrate R with data.world
```{r, echo =TRUE, results='hide'}
saved_cfg <- save_config("eyJhbGciOiJIUzUxMiJ9.eyJzdWIiOiJyLWFuZC1yLXN0dWRpbzpqZXJyaWVrIiwiaXNzIjoiY2xpZW50OnItYW5kLXItc3R1ZGlvOmFnZW50OmplcnJpZWs6OmRlMDQyNGY2LTE3MTctNDg5MS1hNGYyLTYyNTNlOGNhZjEwNSIsImlhdCI6MTU0NzU3MjI3MSwicm9sZSI6WyJ1c2VyX2FwaV9hZG1pbiIsInVzZXJfYXBpX3JlYWQiLCJ1c2VyX2FwaV93cml0ZSJdLCJnZW5lcmFsLXB1cnBvc2UiOnRydWUsInNhbWwiOnt9fQ.GS4kECqsw6IModNmOyYKQf4M_GrOFDM7h9Aj0EGpFDw951en3lxzTnREKInjDnNA08-OMweNhH3LK3vqw6o3Dw")

set_config(saved_cfg)
```
Load data from data.world library
```{r, echo =TRUE, results='hide'}
#link to where the data is housed in this case it is my profile but I could connect to any datasource on data.world
data_link <- "https://data.world/jerriek/homelessness-in-the-us/"
```
Run a query pulling all homeless shelter beds using SQL and create data frame
```{r, echo =TRUE, results='hide'}
total_beds_query <- qry_sql("SELECT  '2007' AS Year, state, total_beds_es_th_total_year_round_beds_es_th
                            FROM 2007
                            UNION
                            SELECT '2008',
                            state, total_beds_es_th_sh_total_year_round_beds_es_th_sh
                            FROM 2008 
                            UNION
                            SELECT '2009',
                            state, total_beds_es_th_sh_total_year_round_beds_es_th_sh
                            FROM 2009 
                            UNION
                            SELECT '2010',
                            state, total_beds_es_th_sh_total_year_round_beds_es_th_sh
                            FROM 2010 
                            UNION
                            SELECT '2011',
                            state, total_beds_es_th_sh_total_year_round_beds_es_th_sh
                            FROM 2011 
                            UNION
                            SELECT '2012',
                            state, total_beds_es_th_sh_total_year_round_beds_es_th_sh
                            FROM 2012 
                            UNION
                            SELECT '2013',
                            state, total_beds_es_th_sh_total_year_round_beds_es_th_rrh_sh
                            FROM 2013 
                            UNION
                            SELECT '2014',
                            state, total_beds_es_th_sh_total_year_round_beds_es_th_sh
                            FROM 2014 
                            UNION
                            SELECT '2015',
                            state, total_beds_es_th_sh_total_year_round_beds_es_th_sh
                            FROM 2015
                            UNION
                            SELECT '2016',
                            state, total_beds_es_th_sh_total_year_round_beds_es_th_sh
                            FROM 2016 
                            UNION
                            SELECT '2017',
                            state, total_beds_es_th_sh_total_year_round_beds_es_th_sh
                            FROM 2017" )


#created dataframe using the query function 
total_beds_df <- query(total_beds_query, dataset = data_link)

#remove all columns with the totals
total_beds_df <- total_beds_df[!total_beds_df$state == "Total", ]
```
Run a query pulling all homeless population data using SQL and create data frame
```{r, echo =TRUE, results='hide'}
total_homelessness_query <- qry_sql("SELECT '2007' AS Year, state, `2007_2`.number_of_cocs AS Number_Of_CoCs, `2007_2`.total_homeless_2007 AS Total_Homeless,  `2007_2`.sheltered_homeless_2007 AS Total_Sheltered_Homeless, `2007_2`.unsheltered_homeless_2007 AS Total_Unsheltered_Homeless
FROM `2007_2`
                                    UNION
                                    SELECT '2008', state, `2008_2`.number_of_cocs, `2008_2`.total_homeless_2008, `2008_2`.sheltered_homeless_2008, `2008_2`.unsheltered_homeless_2008
                                    FROM `2008_2`
                                    UNION
                                    SELECT '2009', state, `2009_2`.number_of_cocs, `2009_2`.total_homeless_2009, `2009_2`.sheltered_homeless_2009, `2009_2`.unsheltered_homeless_2009
                                    FROM `2009_2`
                                    UNION
                                    SELECT '2010', state, `2010_2`.number_of_cocs, `2010_2`.total_homeless_2010, `2010_2`.sheltered_homeless_2010, `2010_2`.unsheltered_homeless_2010
                                    FROM `2010_2`
                                    UNION
                                    SELECT '2011', state, `2011_2`.number_of_cocs, `2011_2`.total_homeless_2011, `2011_2`.sheltered_homeless_2011, `2011_2`.unsheltered_homeless_2011
                                    FROM `2011_2`
                                    UNION
                                    SELECT '2012', state,`2012_2`.number_of_cocs, `2012_2`.total_homeless_2012, `2012_2`.sheltered_homeless_2012, `2012_2`.unsheltered_homeless_2012
                                    FROM `2012_2`
                                    UNION
                                    SELECT '2013', state, `2013_2`.number_of_cocs, `2013_2`.total_homeless_2013, `2013_2`.sheltered_homeless_2013, `2013_2`.unsheltered_homeless_2013
                                    FROM `2013_2`
                                    UNION
                                    SELECT '2014', state, `2014_2`.number_of_cocs, `2014_2`.total_homeless_2014, `2014_2`.sheltered_homeless_2014, `2014_2`.unsheltered_homeless_2014
                                    FROM `2014_2`
                                    UNION
                                    SELECT '2015', state, `2015_2`.number_of_cocs, `2015_2`.total_homeless_2015, `2015_2`.sheltered_homeless_2015, `2015_2`.unsheltered_homeless_2015
                                    FROM `2015_2`
                                    UNION
                                    SELECT '2016', state, `2016_2`.number_of_cocs, `2016_2`.total_homeless_2016, `2016_2`.sheltered_homeless_2016, `2016_2`.unsheltered_homeless_2016
                                    FROM `2016_2`
                                    UNION
                                    SELECT '2017', state, `2017_2`.number_of_cocs, `2017_2`.total_homeless_2017, `2017_2`.sheltered_homeless_2017, `2017_2`.unsheltered_homeless_2017
                                    FROM `2017_2`")

total_homelessness_df <- query(total_homelessness_query, dataset = data_link)
#quick clean-up
total_homelessness_df <- total_homelessness_df[!total_homelessness_df$state == "Total", ]

#remove all rows with no values (NAs)
total_homelessness_df <- na.omit(total_homelessness_df)

#Remove states that will not be used in analysis
unique(total_beds_df$state) #Remove data with GU, PR, MP
unique(total_homelessness_df$state) #Remove data for GU, PR, MP, NA and change KS* to KS

states_remove <- c("GU", "PR", "MP", "NA", "VI")
total_beds_df <- total_beds_df[!total_beds_df$state %in% states_remove, ]
total_homelessness_df <- total_homelessness_df[!total_homelessness_df$state %in% states_remove,]
```
Join the data sets and make additional minor edits
```{r, echo =TRUE, results='hide'}
#Create a joined data frame of the two datasets
joined_data <- left_join(total_beds_df, total_homelessness_df, by = c("Year","state"))

#create a column with full state names
joined_data$state_name <- state.name[match(joined_data$state, state.abb)]
joined_data$region <- state.region[match(joined_data$state, state.abb)]

#fill out DC data
joined_data <- within(joined_data, state_name[state == "DC"] <- "District of Columbia")
joined_data <- within(joined_data, region[state == "DC"] <- "South")

#remove any NA's in joined data
joined_data <- na.omit(joined_data)
```
Upload new joined dataset
```{r, echo =TRUE, results='hide'}
#upload data frame to data.world so that we can play with it in tableau or on data.world
upload_data_frame(file_name = "Joined Data.csv", data_frame = joined_data, dataset = "jerriek/homelessness-in-the-us")
```

